Drought vulnerability of tree growth

### 2024-07-12 scripts/explore_analysis.Rmd
updated code, dag for individual

resistance_div_sp<-ggplot(data = resistance_top4, 
aes(x=resistance.div, 
col=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_density(alpha=0.5) +
  facet_grid(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~yr)+
  scale_color_viridis_d()+ scale_fill_viridis_d()+
  guides(col="none", fill="none")+
  xlim(-5, 5)+
  xlab("inc/long-term increment")+
  geom_vline(xintercept = c(0,1), lty=2)+
  theme_bw()

@@

@@

dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")

resistance <- dendro_inc_clean %>%
  filter(Cno %in% c(5, 15))%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  dplyr::mutate(
    yr=ifelse(Cno==5, 2010, 2015),
        resistance.div = inc_annual/avg_inc,
         resistance.dif = inc_annual-avg_inc,
         resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
         ungroup()%>%
  filter(!is.infinite(resistance.div))%>%
         group_by(Cno)%>%
         #making quantiles for each year
         dplyr::mutate(
         quantiles_div=ntile(resistance.div, 4),
         quantiles_dif=ntile(resistance.dif, 4),
         quantiles_prop=ntile(resistance.prop, 4))

@@

@@

png("growth-precip-thailand/doc/display/resistance_map.png", width=8, height=6, units="in", res=300)
resistance_map
dev.off()
```
```{r basic model}

resistance_all$ciiF<-ordered(resistance_all$cii_min1)

model_2010<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec,
                data=resistance_all[resistance_all$yr==2010,])

@@

model_2015<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec,
                data=resistance_all[resistance_all$yr==2015,])

@@

resistance_model<-resistance_all %>%
  select(resistance.div, calcDBH_min1, ciiF, williams_dec, yr, quantile)

@@

library(lme4)
library(nlme)
model_all<-lm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec + yr,
                data=resistance_model,
                )

model_all
anova(model_all)

@@

>>>>

resistance_div_sp<-ggplot(data = resistance_top4, 
aes(x=resistance.div, 
col=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_density() +
  facet_grid(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~yr)+
  scale_color_viridis_d()+ scale_fill_viridis_d()+
  guides(col="none", fill="none")+
  xlim(-5, 5)+
  xlab("inc/long-term increment")+
  geom_vline(xintercept = c(0,1), lty=2)+
  theme_bw()

@@

```{r associated datasets, echo=F}

#deciduousness
dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")

#habitat data
habitat<-read.csv("growth-precip-thailand/data/HKK-other/HKK_habtype(in).csv")

#plot a tile map of habitat types
ggplot(habitat, aes(x=x, y=y, fill=factor(hab)))+
  geom_tile()+
  scale_fill_viridis_d(option="E", direction=-1)+
  theme_minimal()+
  coord_fixed()

habitat<-habitat %>%
  dplyr::mutate(Col = x/20,
                Row = y/20)

trees<-merge(trees, select(habitat, -c(X, x, y)), by=c("Col", "Row"), all.x=T)

#elevation
#although .rdata, this needs to be read using read.table - thanks Bier!

hkk_elev<-read.table("growth-precip-thailand/data/HKK-other/CTFSElev_hkk.rdata")
png("growth-precip-thailand/doc/display/elevation_resistance_map_2015.png", width=8, height=4, units="in", res=300)
ggplot(hkk_elev, aes(x=col.x, y=col.y, fill=col.elev))+
geom_tile()+
scale_fill_viridis_c(option="E")+
#add contours
geom_contour(aes(z=col.elev), colour="black")+
geom_point(data=resistance_clean %>% filter(yr==2015, sp %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
aes(x=X.x, y=Y, size=as.numeric(DBH.04)), inherit.aes=F, col="black", alpha=0.5)+
facet_wrap(.~sp, nrow=2)+
theme_bw()+
coord_fixed()
dev.off()

#convert to topographic wetness index

library(sf)
library(terra)

#first convert to raster
elev_rast<-rast(hkk_elev$col.elev, nrows=100, ncols=100, xmin=0, xmax=100, ymin=0, ymax=100)
```

@@

resistance <- dendro_inc_clean %>%
  filter(Cno %in% c(5, 15))%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  dplyr::mutate(
    yr=ifelse(Cno==5, 2010, 2015),
        resistance.div = inc_annual/avg_inc,
         resistance.dif = inc_annual-avg_inc,
         resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
         ungroup()%>%
  filter(!is.infinite(resistance.div))%>%
         group_by(Cno)%>%
         #making quantiles for each year
         dplyr::mutate(
         quantiles_div=ntile(resistance.div, 4),
         quantiles_dif=ntile(resistance.dif, 4),
         quantiles_prop=ntile(resistance.prop, 4))%>%
         #remove trees that have -ve long-term increment
          filter(avg_inc>0)

@@

resistance_prop<-ggplot(data = resistance, aes(x=resistance.prop)) +
  geom_density(alpha=0.5) +
  facet_grid(.~yr)+
  xlim(-5, 5)+
  xlab("proportion of change from long-term mean")+
  ggtitle("resistance values for all trees")+
  geom_vline(xintercept = c(-1, 0,1), lty=2)+
  theme_bw()

png("growth-precip-thailand/doc/display/resistance_prop.png", width=6, height=4, units="in", res=300)
resistance_prop
dev.off()

@@

png("growth-precip-thailand/doc/display/resistance_map.png", width=8, height=6, units="in", res=300)
resistance_map
dev.off()
```

```{r resistance with predictor variables, echo=F}

resistance_clean <- resistance %>%
filter(abs(resistance.div)<5)

sum(resistance$resistance.div>5, na.rm=T)/nrow(resistance)

#resistance across species

png("growth-precip-thailand/doc/display/resistance_bysp.png", width=8, height=6, units="in", res=300)
ggplot(data=resistance_clean %>% filter(sp %in% sp_top10), aes(x=factor(sp), y=resistance.prop, fill=factor(sp)))+ 
  geom_violin()+
  geom_boxplot(fill="white", width=0.1)+
  geom_hline(yintercept=c(-1, 0, 1), lty=2)+
  facet_wrap(~yr)+
  ylab("growth decrease from long term mean")+
  #make ylab percentage
  scale_y_continuous(labels=scales::percent)+ 
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))
dev.off()

#resistance with dbh

ggplot()+
  geom_point(data=resistance_clean, aes(x=calcDBH_min1, 
  y=resistance.prop))+
  geom_smooth(data=resistance_clean, aes(x=calcDBH_min1, 
  y=resistance.prop), method="lm")+
  facet_grid(yr~.)+
  theme_bw()

#resistance with cii

ggplot(data=resistance_clean, aes(x=factor(cii_min1), fill=factor(cii_min1),
  #y=resistance.div))+
  y=resistance.prop))+
  geom_violin()+
  geom_boxplot(fill="white", width=0.1)+
  facet_grid(yr~.)+
  theme_bw()

#resistance with deciduousness

ggplot()+
  geom_point(data=resistance_clean, aes(x=williams_dec, 
  #y=resistance.div))+
  y=resistance.prop))+
  geom_smooth(data=resistance_clean, aes(x=williams_dec, 
  #y=resistance.div), method="lm")+
  y=resistance.prop), method="lm")+
  facet_grid(yr~.)+
  theme_bw()

#resistance with habitat

ggplot(data=resistance_clean, aes(x=factor(hab), 
  y=resistance.div, fill=factor(hab)))+
  geom_violin()+
  geom_boxplot(fill="white", width=0.1)+
  scale_fill_viridis_d(option="E", direction=-1)+
  facet_grid(yr~.)+
  theme_bw()

```
```{r basic model}

resistance$ciiF<-ordered(resistance$cii_min1)
resistance$hab<-factor(resistance$hab)

model_2010<-glm(resistance.div ~ calcDBH_min1 + ciiF + 
williams_dec+hab+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec,
                data=resistance[resistance$yr==2010,])

@@

model_2015<-glm(resistance.div ~ calcDBH_min1 + ciiF + 
williams_dec+hab+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec,
                data=resistance[resistance$yr==2015,])

@@

resistance_model<-resistance %>%
  select(resistance.div, calcDBH_min1, ciiF, williams_dec, hab, yr, quantiles_div)

@@

library(lme4)
library(nlme)
model_all<-lm(resistance.div ~ calcDBH_min1 + ciiF + hab+ williams_dec+
                  calcDBH_min1*ciiF + 
                  calcDBH_min1*williams_dec+
                  ciiF*williams_dec + yr,
                data=resistance_model,
                )

summary(model_all)
anova(model_all)

@@

```{r dag, echo=F}

library(dagitty)
library(ggdag)
growth.dag<-dagitty( 'dag {
bb="0,0,1,1"
crowding [pos="0.427,0.650"]
exposure [exposure,pos="0.405,0.325"]
growth [outcome,pos="0.583,0.335"]
size [exposure,pos="0.409,0.151"]
topography [exposure,pos="0.259,0.454"]
water [latent,pos="0.415,0.471"]
exposure -> growth
size -> exposure
topography -> water
water -> crowding
water -> growth
}' )

png("growth-precip-thailand/doc/display/dag_ind.png", width=4, height=4, units="in", res=300)
ggdag_status(growth.dag, text = FALSE, use_labels = "name") +
  guides(color = FALSE) +  # Turn off legend
  theme_dag()
dev.off()

impliedConditionalIndependencies( growth.dag )
```

```{r species DAG, echo=F}

sp.dag<-dagitty('dag {
bb="0,0,1,1"
"canopy position" [exposure,pos="0.276,0.255"]
"median size" [exposure,pos="0.440,0.108"]
deciduousness [exposure,pos="0.458,0.618"]
exposure [latent,pos="0.458,0.238"]
growth [outcome,pos="0.609,0.329"]
water [latent,pos="0.449,0.427"]
"canopy position" -> exposure
"median size" -> exposure
deciduousness -> water
exposure -> growth
water -> growth
}')

png("growth-precip-thailand/doc/display/dag_sp.png", width=4, height=4, units="in", res=300)
ggdag_status(sp.dag, text = FALSE, use_labels = "name") +
  guides(color = FALSE) +  # Turn off legend
  theme_dag()
dev.off()

impliedConditionalIndependencies( sp.dag )
